Human capital drives natural selection in contemporary humans

### 2021-04-21 why-natural-selection.Rmd
More about human capital; simplify graphs; add quadratic regression on n_sibs

* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. This works
    nicely if you throw in a money cost. Done computational example.
* Clarify causality?
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Control for PGS for fertility
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference. 
  - I think we leave this for reviewers.

@@

@@

Since UK Biobank participants are not representative of the wider population,
naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables.

We also check for purifying and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
selection for educational attainment polygenic scores: at lower values of these
scores, the negative effect on fertility is larger (Appendix Figure
\@ref(fig:plot-purifying)).

@@

We emphasize two points. First, this exercise does
*not* apportion the total correlation with fertility between subgroups. 
That is because the polygenic score may also predict the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Second, we do *not* claim that our
observed correlations reflect causal effects of polygenic scores on fertility.
Indeed, polygenic scores correlate with many sources of environmental variation 
that could affect fertility. However, natural selection is a correlational
phenomenon, not a causal one. That is, if a polygenic score correlates with 
low fertility, it is being selected against, whatever the underlying causal
relationship. Differences in equation \@ref(eq:regression) across subgroups can 
be informative about how natural selection works, because they constrain the set 
of possible explanations to those that generate these differences.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category.

@@

These results could be explained by age, if older respondents have lower income
and are less educated, and also show more natural selection on polygenic
scores. However, when we rerun the regressions, interacting the polygenic
score with income category and also with a quadratic in age, the
interaction with income remains significant at 0.05/`r n_regs_ic` for
`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
the PGS with age of leaving full time education and a quadratic in age,
the interaction with age leaving full time education remains significant at
0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
controls are added. The correlation between effect sizes controlling for AFLB,
and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
through two opposing channels: an effect on AFLB, and an opposite-signed effect
on number of children controlling for AFLB. Note that we do not claim that AFLB
is exogenous to an individual's polygenic scores -- indeed, our analyses show
that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
that AFLB mediates part of the relationship between polygenic scores and
fertility, and that once this is controlled for, the remaining part of the
relationship has the opposite sign.

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. One possible explanation for this is
given by the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. If so,
then genetic variants which affect earnings potential in the labour market may
cause opposing effects on fertility. The income effect will lead to natural
selection in favour of earnings-increasing variants. The substitution effect
will do the reverse.

To demonstrate this, we write down a simple model of 
fertility choices. $h$ is an individual's level of human capital, which 
(for now) we identify with his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$, and income $Y\equiv(1-bN)W$. An
individual's payoff is

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in wages (or human capital) 
on this marginal benefit is

@@

The *substitution effect* is negative and reflects that when
wages increase, time devoted to childcare costs more in foregone income.
The positive *income effect* reflects that when wage income is
higher, the marginal loss of wages from children is less painful.
Letting $u(\cdot)$ take the constant relative risk aversion (CRRA)
form $u(y)=\frac{y^{1-\sigma}-1}{1-\sigma}$, where $\sigma>0$ captures
risk aversion in income, we solve for optimal fertility $N^{*}$,
and examine the *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$.
For moderate levels of risk aversion ($0.5<\sigma<1$):

\begin{enumerate}
\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh}<0$.
\item The relationship is weaker (closer to zero) at higher wages and/or
levels of human capital.
\item The relationship is more negative when the time burden of children,
$b$, is larger.
\end{enumerate}

@@

\begin{equation}
U(N_{1},N_{2})=u(Y_{1})+u(Y_{2})+a(N_{1}+N_{2})\label{eq:U}
\end{equation}

Instead of identifying human capital $h$ with wages, we now allow
individuals to choose a level of education $s\in[0,1]$, which has
a time cost in period 1. Education increases period 2 wages, and is
complementary to human capital $h>0$. For period 2 wages we assume
the simple functional form $w(s,h)=sh$. We normalize period 1 wages
to 1. Maintaining the assumption of CRRA utility, we examine total
fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$. For $0.5<\sigma<1$,
facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma=1$).
In addition, for $\sigma$ in a neighborhood of 1:

\begin{enumerate}
\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
is weaker at higher levels of education $s$.
\item The relationship is weaker among those who start fertility in period
2 ($N_{1}^{*}=0$) than among those who start fertility in period
1 ($N_{1}^{*}>0$).
\end{enumerate}

Facts 1-5 match our empirical results. In this model, polygenic scores which
correlate with human capital would have a negative correlation with fertility at
low income and education levels, and a weaker or zero correlation at higher
levels (Figure \@ref(fig:plot-income-educ-level)). For scores which are highly
correlated with human capital, the correlation would also be weaker at higher
levels of the score itself, giving rise to diversifying selection (Appendix
Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
the time burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).

@@

The above does not provide a complete theory of fertility. Rather, it shows that
a relatively simple economic model can explain many of our results. Other
empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
@becker1981treatise.

```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.005) +
        facet_grid(Correlation ~ Controls, labeller = label_both,
                   scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with earnings in a
respondent's first job, and against its correlation with educational attainment,
a predictor of lifetime earnings. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.) The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. Thus, the labour market appears to play an
important role in natural selection. Substitution effects dominate income
effects overall, which fits the known association between income and lower
fertility [@becker1960economic;@jones2006economic].

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
`r n_05_33` out of 33 scores are significant at $p = 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on
fertility.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels. Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$,
this relationship actually becomes stronger at higher wages.

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the risk aversion measure *does* predict the overall number
of children, highly significantly ($p \le 2 \times 10^{-16}$, `r n_risk_sig`/33
regressions). Given that, and the statistical power we get from our large N, we 
believe that the non-result is real: while risk aversion does predict fertility 
in the sample, it is not an important channel for natural selection.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older mothers (22+), selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). These are substantial effects, and since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

@@

>>>>

* Add CIs - done.
* Clarify collider bias/endogeneity - done.
* Formalize theory - done.
  - Comment on results where effects are actually reversed. This works
    nicely if you throw in a money cost. Done computational example.
* Clarify causality?
  - Helps to deal with the "collider bias" argument of reviewer
  - Might also be a better test of mediation via human capital
  - One other possibility. If we use only people with first profession
    teacher, that (as it happens) shuts down the EA3-income correlation.
    It also shuts down the EA3-fertility relationship, suggesting that
    income is indeed the mediator.
* Control for PGS for fertility
* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
  - Point is that we can explain how natural selection is occurring.
* "Discuss empirical predictions for AFLB grounded in dynamic theory"
  - hmm, tricky, maybe we can include it in discussion of economic results at
  end of formal section.
* Why don't women see stronger effects? But note, they do: not significant
  after correction for multiple testing, but a 25% difference. 
  - I think we leave this for reviewers.
* Quadratic regressions for parents - DONE

@@

res_significance <- res_wt_combined %>% 
                      group_by(term) %>% 
                      summarize(
                        all = all(p.value < 0.05/33), 
                        any = any(p.value < 0.05/33)
                      )
res_sig_all <- sum(res_significance$all)
res_sig_any <- sum(res_significance$any)

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

We also check for purifying and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
selection for educational attainment polygenic scores: at higher values of these
scores, the negative effect on fertility is smaller (Appendix Figure
\@ref(fig:plot-purifying)).

@@

We emphasize two points. First, this exercise does
*not* apportion the total correlation with fertility between subgroups. 
That is because the polygenic score may also affect the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Second, we do *not* claim that our
observed correlations reflect causal effects of polygenic scores on fertility.
Indeed, polygenic scores correlate with many sources of environmental variation 
that could affect fertility. However, natural selection is a correlational
phenomenon, not a causal one: if a polygenic score correlates with 
low fertility, it is being selected against, whatever the underlying causal
relationship. Irrespective of causality, differences in equation
\@ref(eq:regression) across subgroups can inform us about how natural
selection works, by constraining the set of possible explanations.

@@

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category.  These results could be
explained by age, if older respondents have lower income and are less educated,
and also show more natural selection on polygenic scores. However, when we rerun
the regressions, interacting the polygenic score with income category and also
with a quadratic in age, the interaction with income remains significant at
$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
if we interact the PGS with age of leaving full time education and a quadratic
in age, the interaction with age leaving full time education remains significant
at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).

@@

To investigate this further, we estimate equation \@ref(eq:regression)
*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
controls are added. The correlation between effect sizes controlling for AFLB,
and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
through two opposing channels: an effect on AFLB, and an opposite-signed effect
on number of children controlling for AFLB. Again, we do not claim that AFLB
is exogenous to an individual's polygenic scores. Indeed, our analyses show
that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
that AFLB mediates part of the relationship between polygenic scores and
fertility, and that once this is controlled for, the remaining part of the
relationship has the opposite sign.

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. A possible explanation for this comes 
from the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect leads higher earners to have more children. The
substitution effect leads them to have fewer children.

Suppose that certain genetic variants correlate with human capital -- skills or 
other characteristics that affect an individual's earnings in the labour market.
These variants may then be associated with opposing effects on fertility. The income
effect will lead to natural selection in favour of earnings-increasing variants
(or variants that are merely associated with higher earnings). The substitution
effect will do the reverse.

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$, and income $Y\equiv(1-bN)W$. An individual's
payoff is

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

The *substitution effect* is negative and reflects that when wages increase,
time devoted to childcare costs more in foregone income. The positive *income
effect* depends on the curvature of the utility function, and reflects that when
income is higher, the marginal loss of income from children is less painful.
Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we solve for optimal fertility $N^{*}$, and examine the
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For moderate levels
of risk aversion ($0.5<\sigma<1$):

\begin{enumerate}
\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
\item The relationship is weaker (closer to zero) at higher wages and/or
levels of human capital.
\item The relationship is more negative when the time burden of children,
$b$, is larger.
\end{enumerate}

@@

\begin{equation}
U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2})\label{eq:U}
\end{equation}

Instead of identifying human capital $h$ with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education increases period 2 wages, and is
complementary to human capital $h > 0$. For period 2 wages we assume
the simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Maintaining the assumption of CRRA utility, we examine total
fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$. For $0.5 < \sigma < 1$,
facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma = 1$).
In addition, for $\sigma$ in a neighborhood of 1:

\begin{enumerate}
\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
is weaker at higher levels of education $s$.
\item The relationship is weaker among those who start fertility in period
2 ($N_{1}^{*} = 0$) than among those who start fertility in period
1 ($N_{1}^{*} > 0$).
\end{enumerate}

Facts 1-5 match our empirical results. In this model, polygenic scores which
correlate with human capital would have a negative correlation with fertility at
low income and education levels, and a weaker or zero correlation at higher
levels (Figures \@ref(fig:plot-income-educ-level)a and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
correlated with human capital, the correlation might also be weaker at higher
levels of the score itself, giving rise to diversifying selection, as we observe
for scores for educational attainment (Appendix
Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
the time burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a relatively simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
@becker1981treatise.

```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.005) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with two measures of
human capital: earnings in a respondent's first job, and educational attainment.
The relationships are strongly negative. Thus, human capital appears to be relevant
to natural selection. Substitution effects dominate income effects overall, which fits the known association between income and lower fertility 
[@becker1960economic;@jones2006economic]. The correlations reverse when we control
for age at first live birth, suggesting that within AFLB categories, the income effect dominates. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.)

@@

Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
controlling for earnings in first job and education levels. Effect sizes are
generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
`r n_05_33` out of 33 scores are significant at $p < 0.05/33$. This suggests that
earnings and education indeed mediate the effect of polygenic scores on
fertility.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note]

[^note]: Indeed, in the one-period model above, while $\frac{dN^{*}}{da} > 0$,
this relationship becomes stronger, not weaker, at higher wages.

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be partially captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of 
income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the risk aversion measure *does* predict the overall number
of children, highly significantly ($p < 2 \times 10^{-16}$, `r n_risk_sig`/33
regressions). Given that, and the statistical power we get from our sample size, 
we believe that the non-result is real: while risk aversion does predict fertility 
in the sample, it is not an important channel for natural selection.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood,
and more lifetime sexual partners. Among older mothers (22+), selection effects
are reversed. Furthermore, the size of selection effects on a polygenic score
correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for these findings.
Because of the substitution effect of earnings on fertility, scores are selected
for when they correlate with low human capital, and this effect is stronger at 
lower levels of income and education.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). Inequalities in polygenic scores are
believed to be important for understanding class mobility and structure 
[see e.g. @belsky2018genetic, @Rimfeld_2018]. If so, these changes are substantial
effects. Since many polygenic scores are predictive of disease risk, they could
also potentially increase health inequalities. In general, the evolutionary
history of anatomically modern humans is related to disease risk
[@benton2021influence]; understanding the role of contemporary natural selection
may aid in mapping the genetic architecture of current health disparities.

@@

We also checked for purifying selection in the parents' generation, using
age/qualification weights multipled by number of siblings. Scores for EA2 and
EA3 again show significant diversifying selection ($p < 0.05/33$, positive coefficient). Other scores including hip circumference and drinks per week 
are not significant.

@@

drake::loadd(res_sibs_quadratic)
sig_sibs <- res_sibs_quadratic %>% 
         filter(grepl("^2", term, fixed = TRUE)) %>% 
         filter(p.value < 0.05/33) %>% 
         pull(score_name)
stopifnot(sig_sibs == c("EA2_noUKB", "EA3_excl_23andMe_UK"))
